Relative Stable Maps in Genus One via Central Alignments

### 2018-02-17 genus-one-relative-maps.tex
changed compatibility into monoid factorisation

\usepackage[english]{babel}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{showlabels}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{marginnote}
\usepackage{stmaryrd}
\usepackage{enumitem}
\usepackage[english]{babel}
\usepackage{yfonts}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{faktor}
\usepackage{xcolor}
\usepackage{xfrac}
\usepackage{tikz,tikz-cd}
\usetikzlibrary{decorations.pathmorphing,decorations.pathreplacing,patterns}

@@

\section{The space of relative centrally aligned maps}
\noindent Recall \cite{RSPW} that the moduli space of maps from centrally aligned curves is obtained by first considering the  Cartesian diagram
\bcd
\tVZc{1}{n}{X}{\beta}\ar[d]\ar[r]\ar[dr,phantom,"\Box"] & \M{1}{n}{X}{\beta}\ar[d] \\
\MM_{1,n}^{\rm{ctr}}\ar[r] & \MM_{1,n}^{\dagger}
\ecd
so that objects of $\tVZc{1}{n}{X}{\beta}$ consist of
\begin{enumerate}
\item a centrally aligned curve $(C,M_C,\delta)$;
\item a stable map $f\colon C \to X$;
\end{enumerate}
subject to the condition that the subcurve $C_0 \subseteq C$, consisting of those components $C_v$ for which $\lambda(v) < \delta$, coincides with the maximal connected genus one subcurve contracted by $f$. We then define
\begin{equation*} \VZc{1}{n}{X}{\beta} \subseteq \tVZc{1}{n}{X}{\beta} \end{equation*}
to be the closed substack consisting of maps satisfying the \emph{factorisation condition}, namely that the map $f\colon C\to X$ factors through the associated contraction to a Smyth curve, i.e. there exists a map $\bar{f}$ making the following square commute:
\bcd
\widetilde C\ar[r]\ar[d] & \overline C\ar[d,"\bar f" left,dotted] \\
C\ar[r,"f"] & X
\ecd
One should think of the factorisation condition as identifying the main component of the moduli space.

\subsection{Definition of the relative space} We now give the central definition of this article.
\begin{dfn}
Fix a vector of tangency conditions $\alpha=(\alpha_1,\ldots,\alpha_n)$ with $\Sigma_{i=1}^n \alpha_i \leq Y \cdot \beta$. Then the \emph{centrally aligned relative space} 
\begin{equation*}\VZc{1}{\alpha}{X|Y}{\beta} \end{equation*}
is defined to be the closed substack of $\VZc{1}{n}{X}{\beta}$ satisfying the following conditions:
\begin{enumerate}
\item \emph{Gathmann's relative condition:} for each connected component $Z$ of $f^{-1}(Y)$, either:
\begin{enumerate}
\item $Z$ is a single point of $C$, in which case if $Z=x_i$ is a marking then we require that the multiplicity of $f$ at $x_i$ along $Y$ is at least $\alpha_i$ (if $Z$ is not a marking then there is no condition);
\item $Z$ is a subcurve $Z \subseteq C$, in which case we require that
\begin{equation*} f^*[Y]|_Z -\sum_{x_i\in Z}\alpha_ix_i\end{equation*}
is an effective class in $\Achow_0(Z)$. Since $Z$ is at most genus one and any line bundle of strictly positive degree on a Gorenstein genus one curve is effective, this condition can be rephrased as the following numerical criterion
\begin{equation*} Y \cdot f_*[Z]+\sum_{j=1}^r m_j\geq \sum_{x_i\in Z}\alpha_i \end{equation*}
where the $m_j$ denote the multiplicities of $f$ along $Y$ at the points $y_j \in Z \cap \overline{C \setminus Z}$ (necessarily a finite number of nodes), together with the requirement that, when we have equality in the above, there is an isomorphism of line bundles:
\begin{equation*} (f|_{Z})^*\OO_X(Y) \otimes \OO_Z\left(\sum_{j=1}^r m_j y_j\right)=\OO_Z\left(\sum_{x_i\in Z}\alpha_ix_i\right)\end{equation*}
\end{enumerate}

\item \emph{Novel condition:} if $v_1,v_2$ are two vertices belonging to $R_\delta$, then the corresponding tangencies to the divisor $m_{j_1}$ and $m_{j_2}$ must be equal. Here $R_\delta$ is the set of vertices $v \in \plC$ with $\lambda(v)=\delta$ equal to the contraction radius.
\end{enumerate}
\end{dfn}

\begin{remark}
Trying to straighten out Gathmann's condition for a curve component of $f^{-1}(Y)$ into an easily verifiable criterion: if $Z$ is a smooth elliptic curve, recall that every line bundle of positive degree on it is effective, hence the relative condition can be rephrased as the following numerical criterion:
\begin{equation*} f_*[Z]\cdot [Y]+\sum_{j=1}^r m^{(j)}\geq \sum_{x_i\in Z}\alpha_i \end{equation*}
together with the additional requirement that, when the above is an equality, there is an isomorphism of line bundles:
\begin{equation}\label{eqn:eq_in_Pic}
(f|_{Z})^*\OO_X(Y) \otimes \OO_Z\left(\sum_{j=1}^r m^{(j)}y_j\right)=\OO_Z\left(\sum_{x_i\in Z}\alpha_ix_i\right) 
\end{equation}

 If $Z$ is reducible, the correct extension is not to impose the same condition componentwise; rather, we should ask the numerical condition for the total degree, and, if $Z$ is obtained by gluing a smooth elliptic curve $E$ with a forest of rational trees $R_i$ at roots $q^E_i$, then the line bundle equality \eqref{eqn:eq_in_Pic} should be required in $\Pic(E)$ after counting each $q^E_i$ towards the left hand side with multiplicity
 \[f_*[R_i]\cdot [Y]+\sum_{\text{external components attached to $R_i$}} m^{(j)}- \sum_{x_j\in R_i}\alpha_j.\]

 That this is the right thing to do follows from the proof that Gathmann's relative space is the projection of Jun Li's moduli space of maps to the expanded pair after contracting the higher levels of the accordion.
\end{remark}

>>>>

\usepackage[english]{babel}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
%\usepackage{showlabels}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{marginnote}
\usepackage{stmaryrd}
\usepackage{enumitem}
\usepackage[english]{babel}
\usepackage{yfonts}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{faktor}
\usepackage{xcolor}
\usepackage{xfrac}
\usepackage{tikz,tikz-cd}
\usetikzlibrary{decorations.pathmorphing,decorations.pathreplacing,patterns}

@@

\section{The space of relative centrally aligned maps}
\noindent Recall \cite{RSPW} that the moduli space of maps from centrally aligned curves is obtained by considering the  Cartesian diagram
\bcd
\tVZc{1}{n}{X}{\beta}\ar[d]\ar[r]\ar[dr,phantom,"\Box"] & \M{1}{n}{X}{\beta}\ar[d] \\
\MM_{1,n}^{\rm{ctr}}\ar[r] & \MM_{1,n}^{\dagger}
\ecd
so that objects of $\tVZc{1}{n}{X}{\beta}$ consist of
\begin{enumerate}
\item a centrally aligned curve $(C,M_C,\delta)$;
\item a stable map $f\colon C \to X$;
\end{enumerate}
subject to the condition that the subcurve $C_0 \subseteq C$, consisting of those components $C_v$ for which $\lambda(v) < \delta$, coincides with the maximal connected genus one subcurve contracted by $f$. They then define
\begin{equation*} \VZc{1}{n}{X}{\beta} \subseteq \tVZc{1}{n}{X}{\beta} \end{equation*}
to be the closed substack consisting of maps satisfying the \emph{factorisation condition}, namely that the map $f\colon C\to X$ factors through the associated contraction to a Smyth curve, i.e. there exists a map $\bar{f}$ making the following square commute:
\bcd
\widetilde C\ar[r]\ar[d] & \overline C\ar[d,"\bar f" left,dotted] \\
C\ar[r,"f"] & X
\ecd
One should think of the factorisation condition as identifying the main component of the moduli space.

\subsection{Definition of the relative space} We now give a definition of the relative space inside the moduli space of centrally aligned stable maps.
\begin{dfn}
Let $X$ be a smooth projective variety with a smooth divisor $Y$. Fix a vector of tangency conditions $\alpha=(\alpha_1,\ldots,\alpha_n)$ with $\Sigma_{i=1}^n \alpha_i = Y \cdot \beta$. Then the \emph{centrally aligned relative space} 
\begin{equation*}\VZc{1}{\alpha}{X|Y}{\beta} \end{equation*}
is defined to be the closed substack of $\VZc{1}{n}{X}{\beta}$ satisfying the following conditions:
\begin{enumerate}
\item \emph{Gathmann's relative condition:} for each connected component $Z$ of $f^{-1}(Y)$, either:
\begin{enumerate}
\item $Z$ is a single point of $C$, in which case if $Z=x_i$ is a marking then we require that the multiplicity of $f$ at $x_i$ along $Y$ is at least $\alpha_i$ (if $Z$ is not a marking then there is no condition);
\item $Z$ is a subcurve $Z \subseteq C$, in which case we require that
\begin{equation*} f^*[Y]|_Z -\sum_{x_i\in Z}\alpha_ix_i\end{equation*}
is an effective class in $\Achow_0(Z)$.
\end{enumerate}

\item \emph{Compatibility condition:} this applies when the minimal subcurve of genus one (the \emph{circuit} or \emph{core}) is contracted to a point in $Y$. In this case we require that the following diagram of solid arrows may be completed to a commutative diagram:
\bcd
Q_{\rm{min}}^{\rm{prestable}}\ar[r]\ar[d] & Q_{\rm{min}}^{\rm{log.map}} \\
Q_{\rm{min}}^{\rm{cen.align}}\ar[dashed,ur]
\ecd
\end{enumerate}
\end{dfn}

\begin{remark}[Explaining Gathmann]
Gathmann's condition is promptly extended to the case that $\Sigma_{i=1}^n \alpha_i \leq Y \cdot \beta$. \textcolor{red}{Question: what is the correct generalisation of the compatibility condition?} Here we try to make the condition for a curve component of $f^{-1}(Y)$ into an easily verifiable criterion: if $Z$ is a smooth elliptic curve, recall that every line bundle of positive degree on it is effective, hence the relative condition can be rephrased as the following numerical criterion:
\begin{equation*} f_*[Z]\cdot [Y]+\sum_{j=1}^r m^{(j)}\geq \sum_{x_i\in Z}\alpha_i \end{equation*}
together with the additional requirement that, when the above is an equality, there is an isomorphism of line bundles:
\begin{equation}\label{eqn:eq_in_Pic}
(f|_{Z})^*\OO_X(Y) \otimes \OO_Z\left(\sum_{j=1}^r m^{(j)}y_j\right)=\OO_Z\left(\sum_{x_i\in Z}\alpha_ix_i\right) 
\end{equation}

 If $Z$ is reducible, the correct extension is not to impose the same condition componentwise; rather, we should ask the numerical condition for the total degree, and, if $Z$ is obtained by gluing a smooth elliptic curve $E$ with a forest of rational trees $T_i$ at roots $q^E_i$, then the line bundle equality \eqref{eqn:eq_in_Pic} should be required in $\Pic(E)$ after counting each $q^E_i$ towards the left hand side with multiplicity
 \[f_*[T_i]\cdot [Y]+\sum_{\text{external components attached to $T_i$}} m^{(j)}- \sum_{x_j\in T_i}\alpha_j.\]

 %That this is the right thing to do follows from the proof that Gathmann's relative space is the projection of Jun Li's moduli space of maps to the expanded pair after contracting the higher levels of the accordion.
\end{remark}

\begin{remark}[Explaining compatibility]
First I wish to define the first line $Q_{\rm{min}}^{\rm{prestable}}\to Q_{\rm{min}}^{\rm{log.map}}$; when looking at a geometric point $\Spec(k=\bar k)$ it is well-known that \[ Q_{\rm{min}}^{\rm{prestable}}=\prod_{q\ \text{node}}\mathbb N_q.\]
Let me assume that all the nodes are internal, i.e. map to $Y$, for all the action happens there, and also that there is only one internal connected component $Z$, which has arithmetic genus one. Then I define $Q_{\rm{min}}^{\rm{log.map}}$ as the saturation of the quotient of
\begin{equation}\label{eq:logminmod} \prod_{q\ \text{node}}\mathbb N_q \times \prod_{S\ \text{internal irr. compo.}}\mathbb N_S\end{equation}
by the relation generated in the groupification of the latter by
\[\langle \rm{Rel}_q=(u_q,\ldots,1,\ldots,-1,\ldots),\ q\ \text{internal node} \rangle\]
where $1$ is in position $S_q^+$ (adjacent to $q$ and further from the circuit), $-1$ in position $S_q^-$ (adjacent to $q$ and closer to the circuit), and $u_q$ in position $q$ is defined as to include the tropical balancing condition (I don't know whether this is standard):
\[u_q=f_*[T_q]\cdot [Y]+\sum_{R_i \text{external compo. attached to }T_q} m^{(i)}- \sum_{x_j\in T_q}\alpha_j,\]
where $T_q$ is the tree of internal components that $q$ separates from the circuit. Notice that the nodes separating $Z$ from the external components have no $1$, so the corresponding relations look like \[(u_q,\ldots,-1,\ldots).\]
The map $Q_{\rm{min}}^{\rm{prestable}}\to Q_{\rm{min}}^{\rm{log.map}}$ is induced by the inclusion of the node factors in \eqref{eq:logminmod}. If we call $e_q$ the image in $Q_{\rm{min}}^{\rm{log.map}}$ of the $q$-th basis vector, then we can verify that the following equations hold: chosen any internal component $S_i$ and any two external component $R_i^1, R_i^2$
such that their path to the circuit goes through $S_i$, then
\begin{equation}
 \sum_{q\in[R_i^1,S_i]}u_qe_q=\sum_{q\in[R_i^2,S_i]}u_qe_q.
\end{equation}
\textcolor{red}{Question: is this monoid factorisation condition a closed one?}
\end{remark}

\begin{remark}
 Here is a tidy example: start with an elliptic curve $E$ contracted to $Y$, with two external components $R_1, R_2$ attached to it and a single marking $x\in E$. The tropical picture would be:
 \begin{figure}
 \begin{tikzpicture}[radius=2pt,scale=0.5]
 \draw[fill=black] (0,0) circle;
 \draw (0,0) -- (4,0);
 \draw[fill=black] (0,2) circle;
 \draw[fill=black] (0,4) circle;
 \draw (0,2) -- (2,3);
 \draw (0,4) -- (2,3);
 \draw (2,3) -- (4,3);
 \draw[fill=white] (2,3) circle;
 \draw[->] (2,1.5) -- (2,.5);
 \end{tikzpicture}
 \end{figure}
The dual picture of the compatibility condition would be:
\begin{tikzpicture}[scale=0.8]
 \draw (0,0) -- (0,2)  (0,0)--(2,0);
 \draw[->] (1,-1.5) -- (1,-.5);
 \draw (0,-4) -- (0,-2) (0,-4) -- (2,-4) (0,-4) -- (2,-2) node[above right]{$e_1=e_2$};
 \draw (1,-2) node[left]{$e_2>e_1$} (2,-3) node[right]{$e_1>e_2$};
 \draw[->] (2.5,1)--(3.5,1);
 \draw[gray!=30] (4,0) -- (4,2)  (4,0)--(6,0);
 \draw (4,0) -- (4.6,2) node[above right]{$m_1/m_2$};
\end{tikzpicture}
which is compatible with the fact that $Q_{\rm{min}}^{\rm{prestable}}\to Q_{\rm{min}}^{\rm{log.map}}$ is given by:
$e_1\mapsto m_2, e_2\mapsto m_1$.
\end{remark}
